Fast, Robust Classification with Applications to Periodic Variable Stars

### 2012-05-10 report/ms.tex
almost done

\section{Introduction}
In this work we design an efficient, robust support vector machine (SVM) algorithm by combining 1) automated tuning of the regularization parameter $\lambda$ and 2) classification of interval data. The performance of this novel SVM is tested using periodic variable star data where features are believed to lie within confidence intervals. The paper begins with a motivating example -- classification of astronomical time series, continues with a discussion of interval SVMs and efficient path algorithms for these models, and concludes with an application of interval SVMs to two data sets from astronomy.

\section{The Data}
\subsection{Background on Periodic Variables}
Modern astronomical surveys observe millions of light sources (stars, galaxies, asteroids) over the course of a mission lasting a few years. Periodic variables, sources which vary periodically in brightness over time, are some of the most interesting. In the 1920's, periodic variables were crucial in Edwin Hubble's discovery the existence of galaxies \cite{berendzen1971hubble}. More recently, periodic variables have played an important role in determining expansion of the universe \cite{freedman2010hubble}.

@@

\section{SVMs for Interval Data}
Support Vector Machines (SVMs) are a popular method for constructing classifiers. Let $x_i \in \mathbb{R}^p$ be the vector of features for observation $i$ and $y_i$ its class (either $-1$ or $+1$). We determine the SVM classifier by solving the optimization problem,
\begin{equation}
\label{eq:svm}
\min_{\beta,\beta_0}  \sum_{i=1}^n (1 - y_i(\beta^Tx + \beta_0))_{+} + \frac{\lambda}{2}\beta^T\beta
\end{equation}
The optimal $\beta^{*}$ and $\beta_0^*$ are used to classify a new observation $x$ using $sign(\beta^{*T}x + \beta_0)$. $\lambda$ is a tuning parameter that balances a tradeoff between maximizing the margin and separating observations in different classes. An optimal $\lambda$ may be chosen using 0-1 loss on a test set or through cross-validation. See \cite{scholkopf2002learning} for an extensive overview of SVMs.

@@

\section{Path Algorithm for Interval SVM}

\subsection{Problem formulation}
Suppose $n$ data points are given: $(x_i,y_i)$ for $i=1,...,n$. Let $\mathcal I_+ = \{i: y_i=+1\}$ and $\mathcal I_- = \{i: y_i=-1\}$ be the set of positive class and negative class respectively. Also let $n_+=|\mathcal I_+|$ and $n_-=|\mathcal I_-|$ be the number of data points in these two classes. In this section, we will derive an efficient algorithm to compute the entire regularization path for the interval SVM problem \eqref{eq:svmInit}.
%% \begin{equation}
%% \label{eq:svmInit}
%% \min_{\beta}\sum_{i=1}^n(1-y_i(\beta_0+\beta^Tx_i)+\rho\sigma_i^T|\beta|)_++\frac{\lambda}{2} ||\beta||_2^2.
%% \end{equation}
%\eqref{eq:q}
%(\ref{eq:q})
Introducing slack variables, the problem becomes:
\begin{align}
\label{eq:primal}
\min & \sum_{i=1}^n\xi_i + \lambda\frac{\beta^T\beta}{2}, \\
\text{subject to } & \xi_i \geq 1 - y_i(\beta_0+\beta^Tx_i)+\rho \sigma_i^Tt,\nonumber \\
                   & \xi_i \geq 0, \text{ for }i=1,2,...,n; \nonumber\\
                   & -t_j\leq \beta_j \leq t_j,\nonumber\\
                   & t_i \geq 0, \text{ for } j=1,2,...,p. \nonumber                      
\end{align}
Let the above problem be the primal problem. The Lagrangian for this problem is
\[
\begin{array}{rl}
L(\xi,\beta_0,\beta,t,\alpha,\gamma,\nu,\mu,c)
%= & \sum_{i=1}^n \xi_i + \lambda \frac{\beta^T\beta}{2} + \sum_{i=1}^n\alpha_i(1 - y_i(\beta_0+\beta^Tx_i)+\rho \sigma_i^Tt-\xi_i) \\
%& - \sum_{j=1}^p\mu_j(t_j-\beta_j) - \sum_{j=1}^p\nu_j(t_j+\beta_j) - \sum_{i=1}^n\gamma_i\xi_i - \sum_{i=1}^nc_jt_j. \\
= & \sum_{i=1}^n(1-\alpha_i-\gamma_i)\xi_i + \lambda\frac{\beta^T\beta}{2} - (\sum_{i=1}^n \alpha_iy_ix_i - (\mu-\nu))^T\beta \\
  & - \sum_{i=1}^n\alpha_iy_i\beta_0 + \sum_{j=1}^p(\rho\alpha_i\sigma_{ji} - \mu_j-\nu_j-c_j)t_j.
\end{array}
\]
Minimizing with respect to the primal variables $(\xi,\beta_0,\beta,t)$, we derive the dual problem as follows:

@@

@@

\begin{center}
\begin{figure}[!h]
   \centering
   \includegraphics[width=0.7\textwidth]{./betaPath.pdf} 
      \caption{The path of $\beta$ for the simulated data.}
   \label{fig:betaPath}
\end{figure}
\end{center}

@@

\section{Application to Variable Star Data Sets}
We now return to the periodic variable classification problem and discuss construction of hyper-rectangles, the specific data sets we use, and results from the models.
\subsection{Constructing the Intervals from the Time Series}
In certain applications, methods for constructing hyper-rectangles (or hyper-spheres) in feature space may be fairly straitforward. For example in El Ghaoui \cite{el2003robust}, the authors discuss applications to micro-array data where several replicates of some experiment are available, and a hyper-rectangle can be chosen to enclose all replicates.

@@

Figure \ref{fig:interval_construction} describes precisely how we subsample the time series and determine interval widths. The details of this procedure could be changed. For example one could bootstrap sample the time measurements, instead of subsampling contiguous sections. Certain computational considerations enter here. It takes around 1 second to derive features for an average length astronomical time series. So sampling, say 20 times, could become prohibitively expensive if the data set is initially at the limits of what is computationally feasible.
\begin{figure*}[ht]
\begin{center}
{\small
\framebox[6in]{
\begin{minipage}[t]{5.8in}
% Insert here your text
\begin{center} \textbf{Constructing Hyper-rectangles from Time Series} \end{center}
\begin{enumerate}
\item Order the light curve measurements in time i.e. $\{(t_1,m_1,e_1),(t_2,m_2,e_2), \ldots , (t_l,m_l,e_l)\}$ where $t_i < t_j$ for $i < j$.
\item Slice each light curve into 5 contiguous sections, producing 5 time series. Slice 1 is $(t_i,m_i,e_i)$ for $0 \leq i \leq .5l$, slice 2 for $.125l \leq i \leq .625l$, slice 3 for $.25l \leq i \leq .75l$ ect.
%% \begin{enumerate}
%% \item $\{(t_1,m_1,e_1), \ldots , (t_{l/2},m_{l/2},e_{l/2})\}$
%% \item $\{(t_1,m_1,e_1),(t_2,m_2,e_2), \ldots , (t_l,m_l,e_l)\}$
%% \end{enumerate}
\item Derive features for each of these slices.
\item Put the interval minimum at the lowest value of the feature for the 5 slices. Put the interval max at the maximum feature value across the 5 slices.
\end{enumerate}
\end{minipage}
}
}
\end{center}
\caption{Description of interval construction algorithm.\label{fig:interval_construction}}
\label{lb}
\end{figure*}

@@

\subsection{Results}

We treat $C$ and $\rho$ as tuning parameters and grid search across them to find the optimal values. We use a grid search for $C$, rather than the path algorithm, because it gives more consistent results. This may be due to a bug somewhere in our path code. In order to assess performance on the grid, we perform 10 fold cross validation, with some initial processing to determine where the grid should be placed.

@@

\begin{figure}[ht]
\begin{center}
\begin{tabular}{cc}
$\begin{array}{cc}
\multicolumn{1}{l}{\mbox{(a)}} & \multicolumn{1}{l}{\mbox{(b)}} \\ \\ \\ [-.35in]
\includegraphics[width=0.45\textwidth]{RR2_svmIntC} &
\includegraphics[width=0.45\textwidth]{CEP_svmIntC}
\end{array}$
\end{tabular}
\end{center}
\caption{(a) Performance across tuning parameters for RR data set. (b) Performance across tuning parameters for CEP data set.}
\label{fig:regularizations}
\end{figure}

>>>>

\section{Introduction}
In this work we design an efficient, robust support vector machine (SVM) algorithm by combining 1) automated tuning of the regularization parameter $\lambda$ and 2) classification of interval data. The performance of this SVM is tested using periodic variable star data where features are believed to lie within confidence intervals. The paper begins with a motivating example -- classification of astronomical time series (Section \ref{sec:data}), continues with a discussion of interval SVMs and efficient path algorithms for these models (Sections \ref{sec:interval_svms} and \ref{sec:path_algorithm}), and concludes with an application of interval SVMs to two data sets from astronomy (Section \ref{sec:application}).

\section{The Data}
\label{sec:data}
\subsection{Background on Periodic Variables}
Modern astronomical surveys observe millions of light sources (stars, galaxies, asteroids) over the course of a mission lasting a few years. Periodic variables, sources which vary periodically in brightness over time, are some of the most interesting. In the 1920's, periodic variables were crucial in Edwin Hubble's discovery the existence of galaxies \cite{berendzen1971hubble}. More recently, periodic variables have played an important role in determining expansion of the universe \cite{freedman2010hubble}.

@@

\section{SVMs for Interval Data}
\label{sec:interval_svms}
Support Vector Machines (SVMs) are a popular method for constructing classifiers. Let $x_i \in \mathbb{R}^p$ be the vector of features for observation $i$ and $y_i$ its class (either $-1$ or $+1$). We determine the SVM classifier by solving the optimization problem,
\begin{equation}
\label{eq:svm}
\min_{\beta,\beta_0}  \sum_{i=1}^n (1 - y_i(\beta^Tx + \beta_0))_{+} + \frac{\lambda}{2}\beta^T\beta
\end{equation}
The optimal $\beta^{*}$ and $\beta_0^*$ are used to classify a new observation $x$ using $sign(\beta^{*T}x + \beta_0)$. $\lambda$ is a tuning parameter that balances a tradeoff between maximizing the margin and separating observations in different classes. An optimal $\lambda$ may be chosen using 0-1 loss on a test set or through cross-validation. See \cite{scholkopf2002learning} for an extensive overview of SVMs.

@@

\section{Path Algorithm for Interval SVM}
\label{sec:path_algorithm}
\subsection{Problem formulation}
Suppose $n$ data points are given: $(x_i,y_i)$ for $i=1,...,n$. Let $\mathcal I_+ = \{i: y_i=+1\}$ and $\mathcal I_- = \{i: y_i=-1\}$ be the set of positive class and negative class respectively. Also let $n_+=|\mathcal I_+|$ and $n_-=|\mathcal I_-|$ be the number of data points in these two classes. In this section, we will derive an efficient algorithm to compute the entire regularization path for the interval SVM problem \eqref{eq:svmInit}.
%% \begin{equation}
%% \label{eq:svmInit}
%% \min_{\beta}\sum_{i=1}^n(1-y_i(\beta_0+\beta^Tx_i)+\rho\sigma_i^T|\beta|)_++\frac{\lambda}{2} ||\beta||_2^2.
%% \end{equation}
%\eqref{eq:q}
%(\ref{eq:q})
Introducing slack variables, the problem becomes:
\begin{align}
\label{eq:primal}
\min & \sum_{i=1}^n\xi_i + \lambda\frac{\beta^T\beta}{2}, \\
\text{subject to } & \xi_i \geq 1 - y_i(\beta_0+\beta^Tx_i)+\rho \sigma_i^Tt,\nonumber \\
                   & \xi_i \geq 0, \text{ for }i=1,2,...,n; \nonumber\\
                   & -t_j\leq \beta_j \leq t_j,\nonumber\\
                   & t_i \geq 0, \text{ for } j=1,2,...,p. \nonumber                      
\end{align}
Let the above problem be the primal problem. The Lagrangian for this problem is
\[
\begin{array}{rl}
L(\xi,\beta_0,\beta,t,\alpha,\gamma,\nu,\mu,c)
%= & \sum_{i=1}^n \xi_i + \lambda \frac{\beta^T\beta}{2} + \sum_{i=1}^n\alpha_i(1 - y_i(\beta_0+\beta^Tx_i)+\rho \sigma_i^Tt-\xi_i) \\
%& - \sum_{j=1}^p\mu_j(t_j-\beta_j) - \sum_{j=1}^p\nu_j(t_j+\beta_j) - \sum_{i=1}^n\gamma_i\xi_i - \sum_{i=1}^nc_jt_j. \\
= & \sum_{i=1}^n(1-\alpha_i-\gamma_i)\xi_i + \lambda\frac{\beta^T\beta}{2} - (\sum_{i=1}^n \alpha_iy_ix_i - (\mu-\nu))^T\beta \\
  & - \sum_{i=1}^n\alpha_iy_i\beta_0 + \sum_{j=1}^p(\rho\alpha_i\sigma_{ji} - \mu_j-\nu_j-c_j)t_j.
\end{array}
\]
Minimizing with respect to the primal variables $(\xi,\beta_0,\beta,t)$, we derive the dual problem as follows:

@@

\begin{center}
\begin{figure}[!h]
   \centering
   \includegraphics[width=0.5\textwidth]{./betaPath.pdf} 
      \caption{The path of $\beta$ for the simulated data.}
   \label{fig:betaPath}
\end{figure}
\end{center}

@@

@@

\section{Application to Variable Star Data Sets}
\label{sec:application}
We now return to the periodic variable classification problem and discuss construction of hyper-rectangles, the specific data sets we use, and results from the models.
\subsection{Constructing the Intervals from the Time Series}
\begin{figure*}[ht]
\begin{center}
{\small
\framebox[6in]{
\begin{minipage}[t]{5.8in}
% Insert here your text
\begin{center} \textbf{Constructing Hyper-rectangles from Time Series} \end{center}
\begin{enumerate}
\item Order the light curve measurements in time i.e. $\{(t_1,m_1,e_1),(t_2,m_2,e_2), \ldots , (t_l,m_l,e_l)\}$ where $t_i < t_j$ for $i < j$.
\item Slice each light curve into 5 contiguous sections, producing 5 time series. Slice 1 is $(t_i,m_i,e_i)$ for $0 \leq i \leq .5l$, slice 2 for $.125l \leq i \leq .625l$, slice 3 for $.25l \leq i \leq .75l$ ect.
%% \begin{enumerate}
%% \item $\{(t_1,m_1,e_1), \ldots , (t_{l/2},m_{l/2},e_{l/2})\}$
%% \item $\{(t_1,m_1,e_1),(t_2,m_2,e_2), \ldots , (t_l,m_l,e_l)\}$
%% \end{enumerate}
\item Derive features for each of these slices.
\item Put the interval minimum at the lowest value of the feature for the 5 slices. Put the interval max at the maximum feature value across the 5 slices.
\end{enumerate}
\end{minipage}
}
}
\end{center}
\caption{Description of interval construction algorithm.\label{fig:interval_construction}}
\label{lb}
\end{figure*}
In certain applications, methods for constructing hyper-rectangles (or hyper-spheres) in feature space may be fairly straitforward. For example in El Ghaoui \cite{el2003robust}, the authors discuss applications to micro-array data where several replicates of some experiment are available, and a hyper-rectangle can be chosen to enclose all replicates.

@@

Figure \ref{fig:interval_construction} describes precisely how we subsample the time series and determine interval widths. The details of this procedure could be changed. For example one could bootstrap sample the time measurements, instead of subsampling contiguous sections. Certain computational considerations enter here. It takes around 1 second to derive features for an average length astronomical time series. So sampling, say 20 times, could become prohibitively expensive if the data set is initially at the limits of what is computationally feasible.

@@

\subsection{Results}
\begin{figure}[ht]
\begin{center}
\begin{tabular}{cc}
$\begin{array}{cc}
\multicolumn{1}{l}{\mbox{(a)}} & \multicolumn{1}{l}{\mbox{(b)}} \\ \\ \\ [-.35in]
\includegraphics[width=0.45\textwidth]{RR2_svmIntC} &
\includegraphics[width=0.45\textwidth]{CEP_svmIntC}
\end{array}$
\end{tabular}
\end{center}
\caption{(a) Performance across tuning parameters for RR data set. (b) Performance across tuning parameters for CEP data set.}
\label{fig:regularizations}
\end{figure}
We treat $C$ and $\rho$ as tuning parameters and grid search across them to find the optimal values. We use a grid search for $C$, rather than the path algorithm, because it gives more consistent results. This may be due to a bug somewhere in our path code. In order to assess performance on the grid, we perform 10 fold cross validation, with some initial processing to determine where the grid should be placed.

@@
